A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-05 latex/main.tex
Updates from Overleaf

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and simple tree representations are therefore necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages. [NOTE: mention overall estimate
of recombination compared with UShER+RIPPLES?]
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

We validated the recombination events in the sc2ts ARG
by running 3SEQ~\cite{Lam2018improved}
on the putative recombinant sequence trios (STAR Methods)\sout{,
and 484 of the 855 (57\%) were classified as recombinant.
Validation status correlates strongly with our QC filter,
with 95\% of the QC-passing recombinants
and 29\% of the QC-failing recombinants
being classified as recombinant.}
.
Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar (\url{https://github.com/phac-nml/rebar}) recombination detection methods
on the putative recombinant sequences,
summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate
at 25\% among the QC-passing recombinants.
This is expected, as CovRecomb is specifically intended
to detect interlineage recombinants
and has previously been shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

>>>>

Millions of SARS-CoV-2 genome sequences were collected in
response to the COVID-19 pandemic, forming a dataset of unprecedented
richness.
Estimated genealogies are fundamental to
gleaning information from this ocean of data, 
and form the primary input to many downstream analyses.
A basic assumption of methods to infer genealogies from viral 
genetic sequence data is that recombination is negligible and 
the genealogy is a tree.
Recombinant lineages have risen to global prevalence in SARS-CoV-2,
and simple tree representations are therefore necessarily
incomplete and potentially misleading.
We present sc2ts, a method to infer reticulate genealogies 
as an Ancestral Recombination Graph (ARG) for SARS-CoV-2
that automatically detects and incorporates recombination.  
We use sc2ts to infer an ARG of 2.48 million genomes,
and show that it captures a broad range of 
known evolutionary features and provides novel 
insights into the relationships between previously identified
recombinant lineages. 
Together, the ARG, inference method and 
suite of open-source analysis tools, 
represent a powerful resource for future research.

@@

We validated the recombination events in the sc2ts ARG
by running 3SEQ~\cite{Lam2018improved}
on the putative recombinant sequence trios (STAR Methods).
Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar (\url{https://github.com/phac-nml/rebar}) recombination detection methods
on the putative recombinant sequences,
summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate
at 25\% among the QC-passing recombinants.
This is expected, as CovRecomb is specifically intended
to detect interlineage recombinants
and has previously been shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.
